`It Is Unbearable To Breath Here': Air Quality, Open Incineration, And Misinformation In Blantyre, Malawi

### 2023-04-06 manuscript/manuscript.tex
Render

\affiliation[1]{organization={ETH Zurich, Department of Mechanical and
Process
Engineering},city={Zurich},country={Switzerland},countrysep={,},postcodesep={}}
\affiliation[2]{organization={Malawi University of Business and Applied
Sciences, Department of Environmental
Health},city={Blantyre},country={Malawi},countrysep={,},postcodesep={}}
\affiliation[3]{organization={University of KwaZulu-Natal, School of
Mathematics, Statistics and Computer
Science},city={Durban},country={South
Africa},countrysep={,},postcodesep={}}

@@

\begin{abstract}
Blantyre, Malawi's Queen Elizabeth Central Hospital (QECH), or Queen's,
as it's known locally, is the country's largest public hospital.
However, Queen's is not served by regular municipal waste collection.
What municipal collection that is done is ad-hoc, sporadic, and based on
the hospital's available financial resources. Rather, most hospital
waste (infectious and non-infectious) is gathered by grounds staff and
openly burned, in several constantly smouldering piles, sending up
clouds of smoke. Speaking directly to an identified knowledge gap on air
quality impacts linked to trash burning and the paucity of African urban
dwellers' voices on air quality issues, this study employed a
mixed-methods approach to both quantitatively measure the air quality
around QECH, and to qualitatively investigate the perceived impacts
amongst staff and caregivers. Low-cost sensors measuring particulate
matter (PM) with particle sizes less than 10μm (PM10) and less than
2.5μm (PM2.5), expressed as the mass of PM per volume of air (μg PMx/m3
air) were recorded every 5 minutes at 8 locations across the QECH for 2
months. Qualitative data collection consisted of 57 interviews with
patients, caregivers and hospital staff (including janitorial and
maintenance staff, nurses, doctors, and administrators). Our results
show that safe air quality thresholds are consistently exceeded across
space and time and that the most problematic air quality surrounds the
shelter for caregivers and those receiving treatment for HIV/AIDs.
Moreover, staff and visitors are severely impacted by the poor air
quality within the space, but feel powerless to make changes or address
complaints. Waste management interventions are desperately needed lest
the patients who arrive at Queen's leave with more health issues than
the ones with which they arrived.
\end{abstract}

@@

\begin{keyword}
    waste management \sep healthcare waste \sep trash burning \sep air
quality \sep Malawi \sep 
    urbanisation
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, boxrule=0pt, enhanced, borderline west={3pt}{0pt}{shadecolor}, breakable, interior hidden, frame hidden]}{\end{tcolorbox}}\fi

@@

Built in 1964, Malawi's year of independence, Queen Elizabeth Hospital
(QECH), or Queen's, as it's known locally, is the country's largest
public hospital. Designed in the Tropical Modernism style of
architecture popular within Africa in the late colonial period, Queen's
open-plan form sprawls across a broad swath of central Blantyre, with
dozens of wards, specialised facilities, and administrative buildings,
linked together by a bewildering maze of covered walkways. A hive of
around-the-clock activity, Queen's bustles from dawn to dusk with a
constant stream of patients, drawn from across the country, and
maintains a city-like feel even after dark as staff and stay-in family
members cook and care for their in-patient dependants. Yet, for the
first-time visitor, it's not the architecture or the crowds that are the
most striking when entering the grounds, but the smell. Immediately
noticeable, even from the road outside the hospital, Queen's is
smothered by an acrid-smelling, white smoke, which hangs over the
grounds day and night: the product of several constantly burning fires
spread across the campus. Breathing this air, when a significant amount
of burning is occurring, primarily at night when the smoke is less
visible, can be incredibly uncomfortable. Moreover, the tropical
architecture, designed to let air and light circulate, means that even
indoors, within patient wards, surgery, and treatment areas, the air
quality can also be unbearable. Poor air quality associated with
burning, through released particulate matter (PM), has been linked to
multiple negative health outcomes like asthma, heart disease, heart
failure, stroke, and cancer among others
\citep{anderson2012clearing, bell2012assessment}. In a space of healing
and recovery, why is such burning occurring, and how can it continue to
persist?

@@

Poor air quality in in African cities can be attributed to multiple
factors, including vehicular emissions, dust, or the use of low-quality
fuels for cooking, lighting, and heating, however, the open burning of
waste has been identified as one of the key contributors to air
pollution within urban areas
\citep{amegah2018proliferation, bulto2020impact}. Although open trashing
burning is a problem in LICs broadly, for instance,
\citet{wiedinmyer2014global} estimate that open trash burning
contributes to 29\% of the total global PM2.5 anthropogenic emissions,
it is of particular concern in Sub-Saharan African (SSA) cities, which
are home to 19 of the world's 50 biggest dumpsites (United Nations
Development Programme, 2021). Particulate matter, or PM, is not specific
to trash burning, but rather the product of any type of incomplete
combustion and is one of the most commonly measured indicators for air
quality. PM is measured in terms of its particle size and generally
classified into PM with particle sizes less than 10μm (PM10) and those
with sizes less than 2.5μm (PM2.5), while one way of measuring the
severity of air pollution is by expressing the mass of PM per volume of
air (μg PMx/m3 air). PM has been linked to multiple negative health
outcomes. Amongst particulate air pollution, PM10 and PM2.5 are of
particular concern as they can penetrate deep into the lungs, and their
exposure has been associated with asthma, heart disease, heart failure,
stroke, and cancer \citep{mohapatra2014effect, xing2016impact}. As a
result, the health impacts of ambient air particle pollution may be
significant, especially for the most vulnerable. For instance, the
\citet{ihme2019ambient} estimates PM related air pollution resulted in
4.14 million deaths worldwide (in 2019), of which, 394,000 occurred in
Africa. Yet, despite its importance, the air quality and health picture
in Africa is worryingly incomplete. For instance, the WHO only collects
data from 10 African countries (covering 39 cities)
\citep{worldhealthorganization2016ambient}. Likewise, robust,
epidemiological data from and for the African context is also lacking,
though emerging data suggests strong correlations between PM and poor
health outcomes \citep{coker2018narrative, heft-neal2018robust}. Moving
forward, it is not just the air at Queen's which lacks clarity.

@@

Due to localised variation (air movement in the immediate area) and the
density of the observations, the plotted 5-minute data obscure
persistent trends. The full set of 5-minute data are plotted and
presented in Appendix A Figure A.1. However, the number of measured
values that exceeded the hazardous limit for both parameters are
summarised in Table~\ref{tbl-peaks}.

@@

{\centering \includegraphics{manuscript_files/figure-pdf/fig-5-minute-intervals-1.pdf}

@@

\caption{\label{fig-5-minute-intervals}PM2.5 and PM10 values collected
every 5 minutes over 10 months at 8 locations.}

>>>>

\affiliation[1]{organization={ETH Zurich, Department of Mechanical and
Process Engineering},addressline={Global Health Engineering,
Clausiusstrasse
37},city={Zurich},country={Switzerland},countrysep={,},postcode={8092},postcodesep={}}
\affiliation[2]{organization={Malawi University of Business and Applied
Sciences, Department of Environmental
Health},city={Blantyre},country={Malawi},countrysep={,},postcodesep={}}
\affiliation[3]{organization={University of KwaZulu-Natal, School of
Mathematics, Statistics and Computer
Science},city={Durban},country={South
Africa},countrysep={,},postcodesep={}}

@@

\begin{abstract}
Blantyre, Malawi's Queen Elizabeth Central Hospital (QECH), or Queen's,
as it's known locally, is the country's largest public hospital.
However, Queen's is not served by regular municipal waste collection.
What municipal collection that is done is ad-hoc, sporadic, and based on
the hospital's available financial resources. Rather, most hospital
waste (infectious and non-infectious) is gathered by grounds staff and
openly burned, in several constantly smouldering piles, sending up
clouds of smoke. Speaking directly to an identified knowledge gap on air
quality impacts linked to trash burning and the paucity of African urban
dwellers' voices on air quality issues, this study employed a
mixed-methods approach to both quantitatively measure the air quality
around QECH, and to qualitatively investigate the perceived impacts
amongst staff and caregivers. Low-cost sensors measuring particulate
matter (PM) with particle sizes less than 10 µm (PM\textsubscript{10})
and less than 2.5 µm (PM2.5), expressed as the mass of PM per volume of
air (μg PMx/m3 air) were recorded every 5 minutes at 8 locations across
the QECH for 2 months. Qualitative data collection consisted of 57
interviews with patients, caregivers and hospital staff (including
janitorial and maintenance staff, nurses, doctors, and administrators).
Our results show that safe air quality thresholds are consistently
exceeded across space and time and that the most problematic air quality
surrounds the shelter for caregivers and those receiving treatment for
HIV/AIDs. Moreover, staff and visitors are severely impacted by the poor
air quality within the space, but feel powerless to make changes or
address complaints. Waste management interventions are desperately
needed lest the patients who arrive at Queen's leave with more health
issues than the ones with which they arrived.
\end{abstract}

@@

\begin{keyword}
    waste management \sep healthcare waste \sep trash burning \sep air
quality \sep Malawi \sep 
    urbanisation
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, interior hidden, sharp corners, enhanced, frame hidden, breakable]}{\end{tcolorbox}}\fi

@@

Built in 1964, Malawi's year of independence, Queen Elizabeth Central
Hospital (QECH), or Queen's, as it's known locally, is the country's
largest public hospital. Designed in the Tropical Modernism style of
architecture popular within Africa in the late colonial period, Queen's
open-plan form sprawls across a broad swath of central Blantyre, with
dozens of wards, specialised facilities, and administrative buildings,
linked together by a bewildering maze of covered walkways. A hive of
around-the-clock activity, Queen's bustles from dawn to dusk with a
constant stream of patients, drawn from across the country, and
maintains a city-like feel even after dark as staff and stay-in family
members cook and care for their in-patient dependants. Yet, for the
first-time visitor, it's not the architecture or the crowds that are the
most striking when entering the grounds, but the smell. Immediately
noticeable, even from the road outside the hospital, Queen's is
smothered by an acrid-smelling, white smoke, which hangs over the
grounds day and night: the product of several constantly burning fires
spread across the campus. Breathing this air, when a significant amount
of burning is occurring, primarily at night when the smoke is less
visible, can be incredibly uncomfortable. Moreover, the tropical
architecture, designed to let air and light circulate, means that even
indoors, within patient wards, surgery, and treatment areas, the air
quality can also be unbearable. Poor air quality associated with
burning, through released particulate matter (PM), has been linked to
multiple negative health outcomes like asthma, heart disease, heart
failure, stroke, and cancer among others
\citep{anderson2012clearing, bell2012assessment}. In a space of healing
and recovery, why is such burning occurring, and how can it continue to
persist?

@@

Poor air quality in in African cities can be attributed to multiple
factors, including vehicular emissions, dust, or the use of low-quality
fuels for cooking, lighting, and heating, however, the open burning of
waste has been identified as one of the key contributors to air
pollution within urban areas
\citep{amegah2018proliferation, bulto2020impact}. Although open trashing
burning is a problem in LICs broadly, for instance,
\citet{wiedinmyer2014global} estimate that open trash burning
contributes to 29\% of the total global PM2.5 anthropogenic emissions,
it is of particular concern in Sub-Saharan African (SSA) cities, which
are home to 19 of the world's 50 biggest dumpsites (United Nations
Development Programme, 2021). Particulate matter, or PM, is not specific
to trash burning, but rather the product of any type of incomplete
combustion and is one of the most commonly measured indicators for air
quality. PM is measured in terms of its particle size and generally
classified into PM with particle sizes less than 10 µm (PM10) and those
with sizes less than 2.5 µm (PM2.5), while one way of measuring the
severity of air pollution is by expressing the mass of PM per volume of
air (μg PMx/m3 air). PM has been linked to multiple negative health
outcomes. Amongst particulate air pollution, PM10 and PM2.5 are of
particular concern as they can penetrate deep into the lungs, and their
exposure has been associated with asthma, heart disease, heart failure,
stroke, and cancer \citep{mohapatra2014effect, xing2016impact}. As a
result, the health impacts of ambient air particle pollution may be
significant, especially for the most vulnerable. For instance, the
\citet{ihme2019ambient} estimates PM related air pollution resulted in
4.14 million deaths worldwide (in 2019), of which, 394,000 occurred in
Africa. Yet, despite its importance, the air quality and health picture
in Africa is worryingly incomplete. For instance, the WHO only collects
data from 10 African countries (covering 39 cities)
\citep{worldhealthorganization2016ambient}. Likewise, robust,
epidemiological data from and for the African context is also lacking,
though emerging data suggests strong correlations between PM and poor
health outcomes \citep{coker2018narrative, heft-neal2018robust}. Moving
forward, it is not just the air at Queen's which lacks clarity.

@@

Due to localised variation (air movement in the immediate area) and the
density of the observations, the plotted 5-minute data obscure
persistent trends. The full set of 5-minute data are plotted and
presented in Appendix A Figure~\ref{fig-five-minute-intervals}. However,
the number of measured values that exceeded the hazardous limit for both
parameters are summarised in Table~\ref{tbl-peaks}.

@@

{\centering \includegraphics{manuscript_files/figure-pdf/fig-five-minute-intervals-1.pdf}

@@

\caption{\label{fig-five-minute-intervals}PM2.5 and PM10 values
collected every 5 minutes over 10 months at 8 locations.}
